---
title: "Day 5: Data Wrangling Practice"
author: "Sara Tyo, Department of Statistics, University of California, Irvine"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: "simple"
    slideNumber: true
    transition: "fade"
    progress: true
    controls: true

execute:
  echo: true
---

# Data Wrangling

## Load Libraries & Data

```{r}
#| code-fold: true
library(tidyverse)
data(diamonds)
```

# dim, colnames, str

```{r}
#| code-fold: true
dim(diamonds)
colnames(diamonds)
str(diamonds)

```

# glimpse, head, head(,10), slice

```{r}
#| code-fold: true
glimpse(diamonds)
head(diamonds)
head(diamonds, 20)
diamonds %>% 
  slice(50:100)

```

# summary

```{r}
#| code-fold: true
summary(diamonds)
```


# subset by row with filter()

- filter diamonds by (1) ideal cut, (2) price > 1000, (3) both, (4) or


```{r}
#| code-fold: true

#?dplyr::filter

diamonds %>% 
  filter(cut == "Ideal")

diamonds %>% 
  filter(price > 1000)

diamonds %>% 
  filter(price > 1000 & cut == "Ideal")

diamonds %>% 
  filter(price > 1000 | cut == "Ideal")
```

# filter for missing values using is.na(), or !(is.na) 

```{r}

children <- c(2,3,4,NA,1)


 # == vs is.na()
is.na(children)
!is.na(children)

```

# subset by column with select()

- subset to have just cut, and color
- reverse order
- use 1:4 to select

```{r}
#| code-fold: true

#?dplyr::select

diamonds %>% 
  select(cut, color)

diamonds %>% 
  select(color, cut)

colnames(diamonds)

diamonds %>% 
  select(3:5)

diamonds %>% 
  select(-cut, -depth)
```

# select helper functions 
- (starts_with, ends_with, -, contains, everything, last_col, last_col offset)

```{r}
#| code-fold: true

diamonds %>% 
  select(starts_with("c"))
diamonds %>% 
  select(ends_with("y"))
diamonds %>% 
  select(contains("c"))
```

# reorder rows with arrange

- maybe want height in ascending/descending order
- makes it easier to find outliers

- numerical: sorts by size/number
- categorical (alphabet) Goes A to B to C ... to Z
- categorical factor: Goes by the storage, so by level

# reorder rows with arrange()
- just color, carat, both, three vars, descending, or descending -.
```{r}
#| code-fold: true

diamonds %>% 
  arrange(color)

diamonds %>% 
  arrange(carat)

diamonds %>% 
  arrange(color, desc(carat))

```

# add or modify columns with mutate()

1. how heavy is each ring's rock
-     need to know how many grams 1 carat is

2. price per carat

3. rename levels cut tolower

4. 

```{r}
#| code-fold: true

#?diamonds

# 1. add grams

copy_diamonds <- diamonds
copy_diamonds <- copy_diamonds %>% 
  mutate("grams" = 0.2 * carats)

colnames(copy_diamonds)


# 2. add price per carat
copy_diamonds <- copy_diamonds %>% 
  mutate("price_per_carat" = price/carat)


# 3. rename/alter columns using mutate [change toupper or tolower]
levels(diamonds$cut)
copy_diamonds <- copy_diamonds %>% 
  mutate(cut = tolower(cut))


# 4. add expensive variable using logic
summary(diamonds$price)
quantile(diamonds$price,.90)
copy_diamonds <- copy_diamonds %>% 
  mutate("expensive" = as.factor(price > 9821)


```

# slice, slice_max, slice_min, slice_sample

```{r}
#| code-fold: true

# ?slice
# ?slice_max

# 1. slice top 10

# 2. slice by max price [most expensive]

copy_diamonds %>% 
  slice_max(price)

# 3. slice by min price [least expensive]

copy_diamonds %>% 
  slice_min(price, n=10)

# 4. slice_sample, some rows (anywhere)
```

# rename, rename(, all_of(key))

```{r}
#| code-fold: true

#?rename

# 1. Rename a column name/label: New then old, maybe change cut to Cut


# 2. Rename multiple columns using a key c(new = "old", new = "old")
# key_1 <- c()
# rename(diamonds,all_of(key_1))

# 3. Rename multiple columns using a key where one var wasn't a part of original
# key_2 <- c(dollars = "price", shade = "color", shiny = "Shiny")
# rename(diamonds, all_of(key_2))
# rename(diamonds, any_of(key_2))

```


# summarise or summarize

- get total num color of F and mean(price)

```{r}
#| code-fold: true

summarise(diamonds,
          "F Color"= sum(color == "F"),
          "mean_price"=mean(price,na.rm=T))
```

# grouped summaries with group_by() and summarize()

# Avg Price and num_cut: Group_by cut

```{r}
#| code-fold: true

copy_diamonds %>% 
  group_by(cut)
  summary("count"=n(), 
          "avg_price"=mean(price), na.rm=T)
```

# Avg Price and num_color: Group_by  color

```{r}
#| code-fold: true

```

# Avg Price and num_cut_color: Group_by cut, color
```{r}
#| code-fold: true

```

# Avg Price and num_expensive: Group_By price > 1000
 -  rename it to expensive
 
```{r}
#| code-fold: true

```

# Pivotting the data

# pivot_wider()

- make long dataset (id each 5 times, test 1- test 5, scores for each test)

```{r}
#| code-fold: true

#?pivot_wider

#long <- data.frame()

# data
# names from
# values_from

#wide <- pivot_wider()
```

# pivot_longer()

- choose columns you want to combine
- what is the new column name
- what is the new column for their values

```{r}
#?pivot_longer

#head(wide)

# data
# cols
# names_to
# values_to

#long <- pivot_longer()
```

# Combine/Merge Data

- wide 1 is some test score data for individuals 1:20
- maybe we have another dataset with their sex, and height that we want to merge

```{r}
#| code-fold: true

wide_2 <- data.frame() # id, sex, height

#?merge

#(x , y, by )

# we can merge it with the long data [diff dimension] by the id

# we can merge it with the wide data [same dimension] by the id
```

## merging by column but the column name is different in each

- maybe instead of both having 'id' as label, maybe the 2nd had 'individual
- 'individual', 'sex', 'age'

```{r}
#| code-fold: true

wide_3 <- data.frame() # same as before but id is now "individual"

#what happens when we run previous, just data, x, and y no by argument 
```

```{r}
#| code-fold: true
# now specify by for x, and by for y [x,y,by.x,by.y]
```

# write new dataset/dataframe and save it

```{r}
#| code-fold: true

# ?write_csv
# ?write_delim
# ?write_excel_csv
# ?write_excel

#write_csv(x = cleaned_dataframe, file = "cleaned_dataframe.csv")
#write_excel_csv(x = cleaned_dataframe, file ="cleaned_dataframe.csv")
#read_csv(file = "cleaned_dataframe.csv")
```


# exporting ggplot images

- explore if carat increase price increases 
- color by the color of diamond
- size by the carat size
- alpha

```{r}
#| code-fold: true


```

# left join, right join, inner_join, full_join

- venn- diagram

```{r}
#| code-fold: true

employees <- data.frame(
  EmployeeID = c(1, 2, 3),
  Name = c("Alice", "Bob", "Charlie"),
  DepartmentID = c(1, 2, NA)
)

departments <- data.frame(
  DepartmentID = c(1, 2, 3),
  DepartmentName = c("HR", "IT", "Finance")
)
glimpse(employees);glimpse(departments)
```

# left join
```{r}
#| code-fold: true
#?left_join
left_join(employees, departments, by = "DepartmentID")
```

# right_join
```{r}
#| code-fold: true

#?right_join
right_join(employees, departments, by = "DepartmentID")
```

# inner_join
```{r}
#| code-fold: true

inner_join(employees, departments, by = "DepartmentID")
```

# full_join
```{r}
#| code-fold: true

full_join(employees, departments, by = "DepartmentID")
```


# bind cols

```{r}
#| code-fold: true

#?bind_cols
#?bind_rows


data_1 <- data.frame(
  id= 1:10,
  sex= c(rep("F", 5), rep("M", 5)),
    x= rnorm(n=10))

data_2 <- data.frame(
  age = sample(size= 10, c(10, 20, 30, 40), replace=T),
  y= rnorm(n= 10))

bind_cols(data_1,data_2)
```

# bind_rows

```{r}
#| code-fold: true

data_3 <- data.frame(
  id= 1:10,
  sex= c(rep("F", 5), rep("M", 5)))

data_4 <- data.frame(
  id= 11:20,
  sex= c(rep("F", 5), rep("M", 5)))

bind_rows(data_3,data_4)
```
